---
title: "EDA of Lacina et al. (2023)"
author: "Daniela Palleschi"
date: "08/23/2024"
format:
  html:
    toc: true
---

This script produces an exploratory data analysis of the online experiment dataset from [@cwiek_novel_2021](https://doi.org/10.1038/s41598-021-89445-4) (OSF: [osf.io/4na58/](https://osf.io/4na58/))

# Set-up

Load required packages.

```{r}
library(tidyverse)
library(here)
library(janitor) # a package for tidying data
```

Load data.

```{r}
df_icon <- read_csv(here("data", "cwiek_2021-online_cleaned.csv")) |> 
  # clean up names
  janitor::clean_names() |> 
  # replace r_t with rt_
  rename_at(vars(contains("l2")), ~ str_replace(., "l2", "L2")) |> 
  rename_at(vars(contains("l1")), ~ str_replace(., "l1", "L1"))
```

# Explore dataset

Check out variables.

```{r}
glimpse(df_icon)
```

```{r}
length(unique(df_icon$id))
```

```{r}
length(unique(df_icon$L1))
```

```{r}
df_icon |> 
  distinct(id, .keep_all = T) |> 
  tabyl(english_L2) |> 
  adorn_totals()
```


There were `r length(unique(df_icon$id))` participants from `r length(unique(df_icon$L1))` L1 groups, 654 of whom speak English as an L2.

```{r}
length(unique(df_icon$unique_item))
```

```{r}
length(unique(df_icon$meaning))
```

There were `r length(unique(df_icon$unique_item))` unique audio files with `r length(unique(df_icon$meaning))` unique meanings.

```{r}
df_icon |> 
  distinct(unique_item, id, .keep_all = T) |> 
  tabyl(english_L2, acc) |> 
  adorn_totals(where = c("col", "row"))
```

There were 4752 unique trials (24 items x 198 participants). Evenly distributed across the four conditions.


```{r}
df_icon |> 
  distinct(id, unique_item, .keep_all = T) |> 
  count(id) |> 
  count(n)
```

We don't have data from all 90 unique items from all participants. The range of responses per participant is 75-90.

## Accuracy distributions

```{r}
df_icon |> 
  distinct(id, unique_item, .keep_all = T) |> 
  tabyl(english_L2, category, acc) |> 
  adorn_totals(where = c("col", "row"))
```

48611 correct responses overall, 26674 incorrect.

# Session Info

```{r}
sessionInfo()
```







