---
title: "Literate programming"
author: "Daniela Palleschi"
date: "08/22/2024"
format:
  html:
    toc: true
---

# Set-up

Set-up environment: load packages and data

## Packages

Install the tidyverse and here packages. This will not run when we render the script.

```{r}
#| eval: false
# install packages
install.packages("tidyverse")
install.packages("here")
```

Load required packages.

```{r}
# load packages
library(tidyverse)
library(here)
```

Print the project directory filepath using the `here` package.

```{r}
# print project directory
here()
```

## Load data

```{r}
df_chromy <- read_csv(here("data", "chromy_et-al_2023_English_final.csv"))
```


# Data exploration

Take a look at the data using `glimpse()` from the `dplyr` package.

```{r}
glimpse(df_chromy)
```

Summary statistics of self-paced reading times (all sentence regions).

```{r}
summary(df_chromy$ReadTimes)
```

Visualise distribution of reading times (milliseconds) for critical sentence regions. I'll use base R functions to keep things simple for now.

```{r}
plot(df_chromy$ReadTimes, df_chromy$WordNo)
```

There were some long observations.

We see some sentence positions had overall shorter reading times (e.g., sentence positions 7 and 11) and some had more spread (e.g., 1 through 6, 12).

# Session Info

```{r}
sessionInfo()
```







