---
title: "Literate programming"
author: "Daniela Palleschi"
date: "08/22/2024"
format:
  html:
    toc: true
---

# Set-up

Set-up environment: load packages and data

## Packages

Install the tidyverse and here packages. This will not run when we render the script.

```{r}
#| eval: false
# install packages
install.packages("tidyverse")
install.packages("here")
```

Load required packages.

```{r}
# load packages
library(tidyverse)
library(here)
```

Print the project directory filepath using the `here` package.

```{r}
# print project directory
here()
```

## Load data

```{r}
df_icon <- read_csv(here("data", "cwiek_2021-online_cleaned.csv"))
```


# Data exploration

Take a look at the data using `glimpse()` from the `dplyr` package.

```{r}
glimpse(df_icon)
```

Summary statistics of longitude and latitude.

```{r}
summary(df_icon$Latitude)
```

```{r}
summary(df_icon$Longitude)
```

Visualise distribution of reading times (milliseconds) for critical sentence regions. I'll use base R functions to keep things simple for now.

```{r}
plot(df_icon$Latitude, df_icon$Longitude)
```

# Session Info

```{r}
sessionInfo()
```







