---
title: "Building a reproducible workflow in R"
subtitle: "Project-oriented workflow"
author: "Daniela Palleschi"
institute: "Humboldt-Universität zu Berlin"
lang: en
date: 2024-10-08
date-format: "ddd MMM D, YYYY"
date-modified: last-modified
language: 
  title-block-published: "Workshop Day 1"
  title-block-modified: "Last Modified"
format: 
  html:
    number-sections: true
    number-depth: 2
    toc: true
    code-overflow: wrap
    code-tools: true
    embed-resources: false
  pdf:
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  revealjs:
    footer: "R workflow"
    output-file: R-workflow-slides.html
editor_options: 
  chunk_output_type: console
bibliography: ../bibs/R-workflow.bib
---

```{r setup, eval = T, echo = F}
knitr::opts_chunk$set(echo = T, # print chunks?
                      eval = T, # run chunks?
                      error = F, # print errors?
                      warning = F, # print warnings?
                      message = F, # print messages?
                      cache = F # cache?; be careful with this!
                      )
```

```{r}
#| echo: false
source(here::here("functions", "print_image.R"))
```


```{r}
#| echo: false
#| eval: false
# run manually
rbbt::bbt_update_bib(here::here("slides", "day1", "R-workflow", "R-workflow.qmd"))
```

# Learning Objectives {.unnumbered .unlisted}

Today we will learn...

-   about reproducibility practices beyond sharing code and data
-   about project-oriented workflows
-   what we will cover in this workshop


# Building a reproducible workflow in R {data-stack-name="R 4 Repro"}

- we now know some important principles of a reproducible workflow
  + and that 'reproducibility' is not black-and-white
  + but even the reproducibility spectrum is an oversimplification [@peng_reproducible_2011]

- some additional resources that provide a list of tips include:
  - @bowers_how_2016; @nagler_coding_1995; @wilson_good_2017; @corker_open_2022

## Broadening the reproducibilty spectrum

-   there are different levels of reproducibility
    -   the *bare minimum* is sharing the code and data
    -   *and* including session information:
        -   which operating system was used
        -   which software/package versions were used
-   going bigger:
    -   project-oriented workflow
    -   project-specific filepaths
    -   contained in a single project folder
-   we will be using RProjects to achieve this

## Project management

-   folder structure
-   project-relative file paths
-   appropriate documentation
    -   e.g., README
-   it's great to map out your project structure early on
    -   but it will grow as you go along
    -   reproducible principles facilitate adapting as it grows

::: {.content-visible when-format="revealjs"}
### Naming conventions {.smaller}
:::
::: {.content-visible unless-format="revealjs"}
### Naming conventions
:::

- there are some "rules" for naming files and folders
    + [The Turing Way: Naming files, folders, and other things](https://the-turing-way.netlify.app/project-design/filenaming.html)
    + [Jenny Bryan: naming things (Reproducible Science Workshop 2015)](https://speakerdeck.com/jennybc/how-to-name-files)

::: columns

::: {.column width="50%"}

1. Avoid special characters
    + ensures machine readability
2. Make names concise but meaningful
    + ensures human-readability
3. Avoid spaces
    + try `CamelCase`, snake case (`snake_case`), or skewer case (`skewer-case`)
    + or use hyphens (`-`) to separate chunks, and underscores (`_`) to connect words of the same chunk

:::

::: {.column width="50%"}
4. Consider default ordering
    + e.g., with dates: `YYYY-MM-DD`
    + with folders or files: numerical prefixes (e.g., `01-data_cleaning.R`, `02-data_visualisation.R`)
5. Be *consistent*

:::

:::

# Data management and sharing {data-stack-name="Data Management"}

## DM(S)P

- Data Management (and Sharing) Plans (DMP) are required by some funding bodies
  + even if not, they're an important 
- questions to consider:
  + do I have data from human participants?
  + do I have data from vulnerable groups (children, patients, etc.)
  + have I collected any identifiable data from humans? (direct or indirect)

### Facilitating data management/sharing

- planning and implementing folder structure, file and variable names
- keep everything relevant to a certain project in one place (i.e., folder)
  + use subfolders appropriately
  + avoid mixing subfolders and files within a single folder

## Literate programming

> Instead of imagining that our main task is to instruct a *computer* what to do, let us concentrate rather on explaining to *human beings* what we want a computer to do.

--- @Knuth_literate_1984, p. 97

::: columns
::: {.column width="100%"}
-   originally used to refer to writing programs
-   but also applies to analysis code
    -   especially if we're aiming for reproducibility
:::

::: {.column width="50%"}
-   main concepts:

    -   code is linear [this pre-dates @Knuth_literate_1984]
    -   informative but concise commenting
:::

::: {.column width="50%"}
-   main benefits:

    -   facilitates maintenance
    -   helpful for future-you, collaborators, etc.
:::
:::

## Documentation

-   metadata
    -   project README
    -   codebook/data dictionary

-   README should contain
    -   a project description
    -   relevant links
    -   description of folder structure

-   can be updated as the project develops

-   README.md files in GitHub/Lab are automatically used as a project description
    -   `.md` is a plain text document
    -   uses markdown syntax

## Version control (not covered in this workshop)

-   git: local tracking
-   useful for the analysis and writing phases
    -   but can be tricky for collaboration
-   GitHub/GitLab: remote tracking
    -   store your changes to your local git repository
    -   then push them to your remote repository
-   safe guards against local hardware/software issues
    -   lost or damaged computer or local files
-   and allows for collaboration or sharing

## Persistant (public) storage

-   GitHub/Lab are sub-optimal
    -   developer-focused
    -   typically lack thorough documentation/metadata
    -   not very user-friendly for non-users
-   OSF, Zenodo
    -   Open Science-focused
    -   can be linked to a GitHub/Lab repository
    -   facilitate thorough documentation
    -   user-friendly

## Writing (not covered in this workshop)

-   dynamic reports with Markdown syntax

    -   e.g., Rmarkdown, Quarto
    -   integration of data, code, and prose
        -   facilitates cross-referencing within document
        -   integration of citation management tools
        -   supports LaTeX syntax for example sentences and tables

-   `papaja` package for APA-formatted Rmarkdown documents

-   challenge: collaboration

    -   not all collaborators know these tools
    -   track changes not currently possible

# Setting up a project

-   required installations/recent versions of:
    -   R
        -   preferably version `4.4.0`, "Puppy Cup" or higher
        -   check current version with `R.version`
        -   download/update: <https://cran.r-project.org/bin/macosx/>
    -   RStudio
        -   preferably a version from 2024
        -   Help \> Check for updates
        -   new install: <https://posit.co/download/rstudio-desktop/>

# Learning objectives 🏁 {.unnumbered .unlisted .uncounted}

Today we learned...

-   about reproducibility practices beyond sharing code and data ✅
-   about project-oriented workflows ✅
-   what we will cover in this workshop ✅

# References {.unlisted .unnumbered visibility="uncounted"}

---
nocite: |
  @seibold_6_nodate
---

::: {#refs custom-style="Bibliography"}
:::
