---
title: "Reproducible analyses in R"
subtitle: "What, Why, and How?"
author: "Daniela Palleschi"
institute: "Leibniz-Zentrum Allgemeine Sprachwissenschaft (ZAS)"
lang: en
date: 2024-10-08
date-format: "ddd MMM D, YYYY"
date-modified: last-modified
language: 
  title-block-published: "Workshop Day 1.1"
  title-block-modified: "Last Modified"
format: 
  html:
    output-file: reproducibility.html
    number-sections: true
    number-depth: 2
    toc: true
    code-overflow: wrap
    code-tools: true
    embed-resources: false
  pdf:
    output-file: reproducibility.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  # clean-revealjs: default
  revealjs:
    output-file: reproducibility-slides.html
    footer: "Open Science Practicies: Reproducibility"
editor_options: 
  chunk_output_type: console
bibliography: ../bibs/reproducibility.bib
---

```{r setup, eval = T, echo = F}
knitr::opts_chunk$set(echo = T, # print chunks?
                      eval = T, # run chunks?
                      error = F, # print errors?
                      warning = F, # print warnings?
                      message = F, # print messages?
                      cache = F # cache?; be careful with this!
                      )
```

```{r}
#| echo: false
source(here::here("functions", "print_image.R"))
```


```{r}
#| echo: false
#| eval: false
# run manually
rbbt::bbt_update_bib(here::here("slides", "day1", "reproducibility.qmd"))
```

# Topics {.unnumbered .unlisted}

-   Open Science Practices
-   Reproducibility: What it is and why/how to practice it
-   Concepts for building a reproducible workflow

# What is Open Science? {data-stack-name="Open Science"}

::: {.fragment}
> “Open science” is an umbrella term used to refer to the concepts of openness, transparency, rigor, reproducibility, replicability, and accumulation of knowledge, which are considered fundamental features of science”

--- @cruwell_seven_2019, p.3
:::

- a movement developed to respond to crisis in scientific research
  + lack of accessibility, transparency, reproducibility, and replicability of previous research
- transparency is key to all facets of Open Science
  + it allows for full evaluation of all stages of science
  
- Open Access, software, data, code, materials...

## Systemic problem in science

- the combination of
  - publication bias
    + journals favour novel, significant findings
  - publish or perish
    + researchers' careers depend on publications

- can/does/did lead to:
  - HARKing
    + Hypothesising After Results are Known
  - p-hacking
    + (re-)running analyses until a significant effect is found
  - replication crisis
    + pervasive failure to replicate previous research
  
## How to practice Open Science

::: columns

::: {.column width="60%"}
- @fig-kathawalla shows some suggestions from @kathawalla_easing_2021

- Open Science is not all-or-nothing
- there are things I consider the bare minimum
  + detailed experiment plan, ideally public
  + openly available materials (e.g., stimuli)
  + share code and data

- the important thing is to do what you can
  


::: {.fragment}
::: {.callout-tip appearance="simple"}
Which Open Science research practices in @fig-kathawalla do you already practice? Are there any you'd like to start implementing?
:::
:::
:::

::: {.column width="40%"}
::: {.fragment fragment-index=1}
```{r echo = F, fig.env = "figure", fig.align = "center", set.cap.width=T, fig.cap="Image source: @kathawalla_easing_2021 (all rights reserved)"}
#| label: fig-kathawalla
#| out-width: 80%
print_image(here::here("media/Kathawalla_research_cycle.png"))
```
:::
:::

:::

::: {.content-hidden when-format="revealjs"}

::: {.callout-note collapse=true}
## The replication crisis

The *replication crisis* refers to the failure of many replication studies to replicate the findings of influential studies. The result of this "crisis" is a move towards Open Science Practices, which emphasise transparency along all stages of research (conception, planning, data collection, data cleaning, data analysis, reporting). 

The issue became more widespread with the publication of @ioannidis_why_2005, entitled *Most published research findigns are false*. This paper defined bias in terms of design, analysis, and presentation factors with a focus on issues with *p*-values and statistical power. Since then there have been many replication attempts of influential (and lesser influential) papers. Strikingly, @open_science_collaboration_estimating_2015 reports 100 psychological studies run by 270 collaborators. They reported significant effects in only 36% of replications, with 47% of originally reported effects falling within 95% CIs of the replication effect. In essence: fewer significant findings and smaller effect sizes were found in replication studies compared to the original 100 studies [@fig-replication].

```{r}
#| echo: false
#| out-width: "60%"
#| label: fig-replication
#| fig-cap: "Source: @open_science_collaboration_estimating_2015 (all rights reserved)"
print_image(here::here("media/Open_Sci_collab_2015_replication-effects.png"))
```
:::

:::

# Reproducibility {data-stack-name="Reproducibility"}

## *What* is reproducibility?

-   one piece of the Open Science pie
-   generating the same results with the same data and analysis scripts
  -   seems obvious, but requires organisation and forethought before and during data collection/analysis
-   bare minimum: share the code and the data [@laurinavichyute_share_2022]

### Reproducibility vs. replication

- the two terms have been used interchangably in the past [e.g., in the title of @open_science_collaboration_estimating_2015]
  + we'll define them as follows (and this is becoming the standard distinction, imo)

::: columns

::: {.column style="width: 50%; font-size: 60%; color: #8D203D"}

**Reproducibility**

- re-analysing the same data using (ideally) the same scripts, software...
- aim: produce the same results (means, model estimates, etc.)
- why: tests for errors, coding mistakes, biases, etc.
:::

::: {.column style="width: 50%; font-size: 60%; color: #8D203D"}

**Replication**

- re-running a previous experiment, ideally with the same materials, set-up...
  - ideally the same analysis workflow as the original study (i.e., like reproducing the analyses but with new data)
- aim: test whether results are replicated with new data in terms of direction and magnitude

:::

:::

- in short:
  - reproducibility = re-analysis of the *same data*
  - replication = collection of *new data*

## *Why* implement reproducibility in my workflow?

- firstly: the help future you (or collaborators/other researchers)!
  + you may return to your analyses tomorrow, next month, or next year
- to ensure robustness and to document your steps
  + 'researcher degrees of freedom' and the 'garden of forking paths': there's more than one way to analyse a certain dataset
  + we can try to plan ahead in detail (e.g., pre-reigster your analysis plan), but there will always be decisions made that were not foreseen
- lastly: it makes your life *much* easier and streamlines your workflow

## *How* to implement reproducibility?

- not exactly straightforward
  + there are degrees of reproducibility
  + the rest of our time will be spent on this topic
- sharing code and data is a first step
  + think of the FAIR principles of data sharing
  + apply them to sharing analyses as well

### Practice FAIR principles

::: {.column width="100%"}
-   guidelines for sharing digital resources
-   refers broadly to (meta)data, let's extend them to analysis code
:::

::: columns
::: {.column width="40%"}
```{r}
#| echo: false
#| out-width: "80%"
#| fig-align: center
#| label: fig-FAIR
#| fig-cap: "Source: [National Library of Medicine](https://www.nlm.nih.gov/oet/ed/cde/tutorial/02-300.html) (all rights reserved)"
print_image(here::here("media/FAIR_nlm_nih.png"))
```
:::

::: {.column width="60%"}
-   **findable** and **accessible**: where materials are stored
    -   in *findable* repositories
    -   that are *accessible*, i.e., do not require an account
-   **interoperable** and **reusable**: format of data (and code)
    -   the importance of future use
    -   and use beyond your precise computational environment
:::
:::

### Conduct a code review

-   a great way to test the FAIR principles
    -   code review!
    -   i.e., have a colleague try to access your data/run your code
        -   either via an online repository
        -   or send them your project folder

### The reproducibility spectrum

- reproducibility is on a continuum, referred to as the *reproducibility spectrum* in @peng_reproducible_2011 (@fig-peng_2011)
  + *linked* means "*all data, metadata, and code [is] stored and linked with each other and with corresponding publications*" [@peng_reproducible_2011, p. 1227]
  + *executable* is not explained, and is more difficult to guarantee long-term as it depends on software versions
  + but at minimum we can assume it refers to code running on someone else's machine

::: {.fragment fragment-index=1}
```{r}
#| echo: false
#| out-width: "50%"
#| fig-align: center
#| label: fig-peng_2011
#| fig-cap: "Source: @peng_reproducible_2011"
print_image(here::here("media/peng_2011_reproducibility_spectrum.png"))
```
:::
### Share the code, not just the data

-   Why?
    + key details are often missing from 'Methods' sections
-   suggestions for researchers from @laurinavichyute_share_2022

::: columns
::: {.column width="50%"}
1.  Share data in usable form
    -   with pre-processing code
2.  Use publicly accessible repositories
    -   e.g., OSF
3.  Use non-proprietary data formats
    -   e.g., not `.xls` files (Excel)
:::

::: {.column width="50%"}
4.  Provide documentation
    -   e.g., README, data dictionaries
5.  Share code *and* data
    -   they estimate a 38% increase in reproducibility
6.  Teach data management and computing skills
    -   that's what this workshop is for!
:::
:::

### Data and code $\neq$ Reproducibility

-   access to data and code does not mean analyses are reproducible

-   what can go wrong? Examples from @laurinavichyute_share_2022

::: columns
::: {.column width="50%"}
1.  Data problems
    -   inaccessible data
    -   incomplete data (e.g., 2/3 experiments)
2.  Code problems
    -   incomplete code
    -   error messages
    -   code rot: outdated syntax or environment
    -   proprietary software
:::

::: {.column width="50%"}
3.  Documentation problems
    -   data difficult to interpret
    -   no README file/data dictionary
    -   unclear folder/file/variable naming convention
    -   manuscript contradicts code
4.  Unclear terms of use
    -   no licence specification
:::
:::


### What should (ideally) be shared?

::: columns
::: {.column width="50%"}
-   materials
    -   protocols
    -   stimuli
    -   experiment set-up
-   documentation
    -   README
    -   metadata
:::

::: {.column width="50%"}
-   data
    -   raw
        -   e.g., text files, audio, video, or images
    -   processed
-   analysis code
    -   pre-processing
    -   analyses
:::
:::

::: columns
::: {.column width="50%"}
-   materials are helpful for replication
    -   but also for inspection of e.g., design
:::

::: {.column width="50%"}
-   data and code are necessary for reproducibility
    -   along with proper documentation of software used
:::
:::

# Reproducibility rates of published works {data-stack-name="Reproducibility rates"}

- rates of reproducibility vary across fields [see @bochynska_reproducible_2023 for a review]
    - open access: 25-65%
    - data and analyses sharing: 11-33%
    - pre-registrations: 0-3%

## Reproducibility rates in linguistic research {.smaller}

::: columns
::: {.column width="60%"}
```{r}
#| echo: false
#| out-width: "95%"
#| fig-align: center
#| label: fig-bochynska
#| fig-cap: "Source: @bochynska_reproducible_2023, p. 11 (all rights reserved)"
print_image(here::here("media/bochynska_reproducibility.png"))
```
:::

::: {.column width="40%"}
-   meta-analysis of 519 randomly sampled articles from various linguistic journales
    -   pre- and post-reproducibility crisis (2008/9, 2018/19) [@bochynska_reproducible_2023]
    -   differentiated between primary (collected for study) and secondary (pre-existing) data
-   reported a post-RC increase in shared materials, data, and analyses
    -   but still low rates of each
-   higher rates of secondary data sharing, presumably due to publicly available corpora
-   data shared more often than analyses, pre- and post-RC
:::
:::

## Journal of Memory and Language

-   meta-analysis of articles from JML [@laurinavichyute_share_2022]
    -   before and after an Open Science Policy was introduced in 2019

```{r}
#| echo: false
#| out-width: "95%"
#| fig-align: center
#| label: fig-laurinavichyute
#| fig-cap: "Source: @laurinavichyute_share_2022, p. 5 (all rights reserved)"
print_image(here::here("media/laurinavichyute_reproducibility.png"))
```

-   code and data availability improved
-   but reproducibility rate ranged from 34-56%, depending on criteria
-   higher rates compared to field-wide meta-analysis [@bochynska_reproducible_2023]

# Steps we'll take

1. Open source software:
    - R, an open source statistical programming language
    - in RStudio, an IDE (integrated developer environment)
    - with [R Projects](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects)

2. Project-oriented workflow: 
    - establish folder structure
    - and file/variable naming conventions
    - use project-relative filepaths with the `here` package
    - establish and maintain project-relative package library with `renv` (time permitting)
  
::: {.content-visible when-format="revealjs"}
# {-}
:::
    
3. Practice literate programming:
    - writing clean, commented, linear code
    - in dynamic reports (e.g., Quarto, R markdown)
    - practice modularity, i.e., 1 script = 1 purpose

4. Sharing and checking our code
    - uploading our code and data to an OSF repository
    - conducting a code review

# Topics 🏁 {.unnumbered .unlisted .uncounted}

-   Open Science Practices ✅
-   Reproducibility: What it is and why/how to practice it ✅
-   Concepts for building a reproducible workflow ✅

# References {.unlisted .unnumbered visibility="uncounted"}

---
nocite: |
---

::: {#refs custom-style="Bibliography"}
:::
