---
title: "R-Projects"
subtitle: "Creating a project-oriented workflow in R"
author: "Daniela Palleschi"
institute: Leibniz-Zentrum Allgemeine Sprachwissenschaft
lang: en
date: 2024-10-08
date-format: "ddd MMM D, YYYY"
date-modified: last-modified
language: 
  title-block-published: "Workshop Day 1"
  title-block-modified: "Last Modified"
format: 
  html:
    output-file: R-Projects.html
    number-sections: false
    toc: true
    code-overflow: wrap
    code-tools: true
    embed-resources: false
  pdf:
    output-file: R-Projects.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  revealjs:
    footer: "R-Projects and {here}"
    output-file: R-Projects-slides.html
editor_options: 
  chunk_output_type: console
bibliography: ../bibs/RProjects.bib
execute:
  echo: false
---

```{r}
#| echo: false
#| eval: false
rbbt::bbt_update_bib(here::here("pages", "RProjects.qmd"))
```

# Learning Objectives {.unnumbered .unlisted}

Today we will...

-   learn about project-oriented workflows
-   create an R-Project
-   use project-relative filepaths with the `here` package

# Installation requirements

-   required installations/recent versions of:
    -   R
        -   at least version `4.4.0`, "Puppy Cup"
        -   check current version with `R.version`
        -   download/update: <https://cran.r-project.org/bin/macosx/>
    -   RStudio
        -   at least version `2023.12.1.402`, "Ocean Storm"
        -   Help \> Check for updates
        -   new install: <https://posit.co/download/rstudio-desktop/>

# Project-oriented workflow  {data-stack-name="Project-oriented workflow"}

1. Folder structure:
    + keeping everything related to a project in one place
    + i.e., contained in a single folder, with subfolders as needed
2. Project-relative working directory
    + the project folder should act as your working directory
    + all file paths should be relative to this folder

## Folder structure

- a core computer literacy skill
  + keep your Desktop as empty as possible
  + have a sensible folder structure
  + avoid mixing subfolders and files
    + i.e., if a folder contains subfolders, ideally it should not contain files

# R-Projects  {data-stack-name="R-Projects"}

- in data analysis, using an IDE is beneficial
  + e.g., RStudio
- most IDEs have their own implementation of a Project
- in RStudio, this is the R-Project
  + creates a `.Rproj` file in a project folder
  + stores project settings
- you can have several R-Projects open simultaneously
  + and run several scripts across projects simultaneously
- most importantly, R-Projects (can) centralise a specific project's workflow and file path
- to read more about R-Projects, check out [Section 6.2: Projects](https://r4ds.hadley.nz/workflow-scripts.html#projects) from @wickham_r_2023 [or [Ch. 8 -  Workflow: Projects](https://r4ds.had.co.nz/workflow-projects.html) in @wickham_r_2016]

## Creating a new Project

- when?
  + whenever you're starting a new course oR-Project which will use R
- why?
  + to keep all the relavent materials in one place
- where?
  + somewhere that makes sense, e.g., a folder called `SoSe2024` or `Mastersarbeit`
- how?
  + `File > New Project > New Directory > New Project > [Directory name] > Create Project`

### {.unnumbered .unlisted}

::: {.callout-tip}
# New R-Project

Create a new R-Project for this workshop

  + `File > New Project > New Directory > New Project > [Directory name] > Create Project`
  + make sure you choose a sensible location
:::

## Opening a Project

- to open a project, locate its `.Rproj` file and double-click
- or if you're already in RStudio, you can use the `Project (None)` drop-down (top right)

:::: {.columns}

::: {.column width="50%"}

```{r}
#| label: fig-click-open
#| fig-cap: Double-click `.Rproj`
#| out-width: "80%"
magick::image_read(here::here("media", "rstudio_click_open.png"))
```

:::

::: {.column width="50%"}

```{r}
#| label: fig-project-open
#| fig-cap: Open from RStudio
#| out-width: "80%"
magick::image_read(here::here("media", "rstudio_project_open.png"))
```

:::


::::

## Adding a README file

- `File > New File > Markdown File` (*not* R Markdown!)
  + add some text describing the purpose of this project
  + include your name, the date
  + use Markdown formatting (e.g., `#` for headings, `*italics*`, `**bold**`)
- save as `README.md` in youR-Project directory

## Global RStudio options

:::: {.columns}

::: {.column width="50%"}

```{r}
#| label: fig-rstudio-settings
#| fig-cap: RStudio settings for reproducibility
#| out-width: "80%"
magick::image_read(here::here("media", "RStudio_global-options.png"))
```

:::

::: {.column width="50%"}

- `Tools > Global Options`
  + **Workspace**: Restore .RData into workspace at startup: NO
  + Save workspace to .RData on exit: Never

- this will ensure that you are always starting with a clean slate
  + and that your code is not dependent on some pacakge or object you created in another session
- this is also how RMarkdown and Quarto scripts run
  + they start with an empty environment and run the script linearly

:::


::::

## {.unnumbered .unlisted}

::: {.callout-tip}
## Global settings

Change your Global Options so that

  + **Workspace**: Restore .RData into workspace at startup: NO
  + Save workspace to .RData on exit: Never
:::

## Identifying your R-Project {.smaller}

- there are a ways to check which (if any) R-Project you're in
  + there are 6 differences between @fig-noproject and @fig-project
  + which is in an R-Project session?

::: {.panel-tabset}

### Spot the differences

:::: {.columns}

::: {.column width="45%"}

```{r}
#| label: fig-noproject
#| fig-cap: RStudio Session A
#| out-width: "100%"
magick::image_read(here::here("media", "rstudio_noproject.png"))
```

:::

::: {.column width="5%"}
:::

::: {.column width="45%"}

```{r}
#| label: fig-project
#| fig-cap: RStudio Session B
#| out-width: "100%"
magick::image_read(here::here("media", "rstudio_project.png"))
```

:::

::::

### Show the differences

```{r}
#| label: fig-project-diffs
#| fig-cap: How to tell if you're in a project
#| out-width: "80%"
magick::image_read(here::here("media", "RProject_spot-the-diffs.png"))
```

:::

# Folder structure  {data-stack-name="Folder structure"}

- some folders you'll typically want to have:
  + `data`: containing your dataset(s)
  + `scripts` (or `analyses`, etc.): containing any analysis scripts
  + `manuscript`: containing any write-ups of your results
  + `materials`: containing relevant experiment materials (e.g., stimuli)
- let's just create the first 2 (`data` and `scripts`)

### `data/`

- do you have "raw", i.e., pre-processed data?
  + if so, you might want to create a `raw` sub-folder
  + and any other relevant sub-folders (e.g., `processed` or `tidy`)
- download [the dataset](https://osf.io/ushpw) from the workshop repo [from @cwiek_novel_2021]
  + *or*, move a dataset of your own to this folder
  
> Participants listened to each of the 90 vocalizations, and for each one, guessed its meaning from 6 alternatives (the correct meaning plus five randomly generated alternatives).
>
-- @cwiek_novel_2021

```{r}
#| eval: false
write.csv(file = "data/cwieck_2021-online_cleaned.csv", "https://raw.githubusercontent.com/bodowinter/iconicity_challenge/refs/heads/master/data/online_cleaned.csv")
```

```{r}
#| echo: false
#| eval: false
write.csv(file = here::here("data/cwieck_2021-online_cleaned.csv"), "https://raw.githubusercontent.com/bodowinter/iconicity_challenge/refs/heads/master/data/online_cleaned.csv")
```


### `scripts/`

- try to create a single script for each "product"
  + e.g., anonymised data, 'cleaned' data, data exploration, visualisation, analyses, etc.
- you can create sub-folders as the project develops and move scripts around
  + for now, let's create a new script to take a look at our data

### {.unnumbered .unlisted}

::: {.callout-tip}
## New script

Create a new script:

1. `File > New File >` Choose your preferred script type
5. Save it in your `scripts/` folder: `File > Save as...`
:::

### Load in the data

- load in the data however you normally would
  + e.g., `read.csv()`, `readr::read_csv()`, ...
  
::: {.content-hidden when-format="revealjs"}
### Exercise: mini-Code Review
:::

::: {.callout-tip}
#### R-Project template

::: nonincremental
1. Download the R-Project template at [https://osf.io/ctmwj/](https://osf.io/ctmwj/)
2. Open (or switch to) `rproject-template.Rproj`
3. Inspect the folder structure and the files.
4. Look at the `scripts/` folder. Is it clear which scripts should be run first?
5. Try running `02-visualisation.R` first. Do you encounter any problems?

:::
:::

# `here`-package {data-stack-name="{here}"}

- `here` package [@here-package] enables file referencing
  + avoids the use of `setwd()`

::: {.content-visible when-format="revealjs"}
## {.unnumbered .unlisted}
:::
```{r}
#| label: fig-here
#| fig-cap: Illustration by [Allison Horst](https://github.com/allisonhorst)
magick::image_read(here::here("media", "Horst_here.png"))
```


## The problem with `setwd()`

::: {.fragment}

> If the first line of your R script is
>
> `setwd("C:\Users\jenny\path\that\only\I\have")`
>
> I will come into your office and SET YOUR COMPUTER ON FIREðŸ”¥.

--- [Jenny Bryan](https://x.com/hadleywickham/status/940021008764846080)

:::

- `setwd()` depends on your entire machine's folder structure
- `setwd()` breaks when you
  + send youR-Project folder to a collaborator
  + make your analyses open
  + change the location of youR-Project folder
- using slashes is also dependent on your operating system

::: {.content-visible when-format="revealjs"}
### {.unnumbered}
:::

- trying to use somebody else's (or your former) folder path will result in a warning message like:

::: {.fragment}
`Error in setwd("/Users/danielapalleschi/Documents/R/rproject-template") : `
 ` cannot change working directory`
:::

## The benefit of `here()`

- uses the top-level directory of your Project as the working directory
  + meaning we never need to specify the path to our project folder relative to our current higher-level folder structure
- can separate folder names with a comma
  + meaning it doesn't matter if the original code was written on a Mac or a Windows machine

## {.unlisted .unnumbered}

::: {.callout-tip}
# `here`

In your R Project, load the `cwiek_2021-` using `here`

1. Install `here` (e.g., `install.packages("here")`)
2. Load `here` at the beginning of your package
    + or use `here::` before calling a function
3. Use the `here()` function to load in your data
4. Inspect the dataset however you usually would (e.g., `summary()`, `names()`, etc.)
4. Save your script

:::

## `here::here()`

- install package

```{r filename = "In the Console"}
#| eval: false
#| echo: true
install.packages("here")
```

- load package and call the `here` function

```{r}
#| eval: false
#| echo: true
# load package
library(here)

# read in data
df_data <- read.csv(here("data", "data_lifetime_pilot.csv"))
```

- or directly call the `here` function without loading the package

```{r}
#| eval: false
#| echo: true

# read in data without loading here
df_data <- read.csv(here::here("data", "data_lifetime_pilot.csv"))
```

- note that I stored the data with the prefix `df_`
  + `df` stands for dataframe
- I recommend using object-type defining prefixes for all objects in your Environment
  + e.g., `fit_` for models, `fig_` for figures, `sum_` for summaries, `tbl_` for tables, etc.

## {.unlisted .unnumbered}

::: {.callout-tip}
# Reproduce your analysis

1. Perform some data exploration (e.g., with `names()`, `summary()`, `dplyr::glimpse()`, whatever you typically do)
1. Save your script, then close RStudio/your R-Project.
2. Re-open the project. Can you re-run the script?
:::

# Learning objectives ðŸ {.unnumbered .unlisted .uncounted}

Today we learned...

-   learn about project-oriented workflows âœ…
-   create an R-Project âœ…
-   establish a self-contained project environment with `here` âœ…

# References {.unlisted .unnumbered visibility="uncounted"}

---
nocite: |
  @bryan_what_nodate
  @bryan_chapter_nodate
  @noauthor_using_2024
---

::: {#refs custom-style="Bibliography"}
:::
